---
title: "Week 05 - Exploratory Data Analysis (EDA)"
format: html
---

## Overview

This week we continue our exploration of EDA, focusing on **distributions**, **outlier detection**, and **data transformation**. By the end of this week you should be able to:

- Describe the shape, center, and spread of a distribution
- Identify and handle outliers appropriately
- Apply transformations to skewed data
- Summarize grouped data using `dplyr`

---

## 1. Recap: What is EDA?

Exploratory Data Analysis is the process of using statistical and visual tools to understand the structure of your data **before** modeling. It helps you:

- Detect errors or anomalies
- Understand variable distributions
- Discover relationships between variables
- Formulate hypotheses

---

## 2. Describing Distributions

A distribution describes how values of a variable are spread. Key characteristics:

- **Shape**: symmetric, left-skewed, right-skewed, bimodal
- **Center**: mean, median
- **Spread**: standard deviation, IQR, range
- **Outliers**: unusual observations far from the bulk of the data

### 2.1 Measures of Center and Spread

```{r}
# Load libraries
library(tidyverse)

# Use the built-in mpg dataset
data(mpg)

# Summary statistics
summary(mpg$hwy)
```

```{r}
# Manual calculations
mean_hwy   <- mean(mpg$hwy)
median_hwy <- median(mpg$hwy)
sd_hwy     <- sd(mpg$hwy)
iqr_hwy    <- IQR(mpg$hwy)

cat("Mean:   ", mean_hwy, "\n")
cat("Median: ", median_hwy, "\n")
cat("SD:     ", sd_hwy, "\n")
cat("IQR:    ", iqr_hwy, "\n")
```

> **Note:** When the mean > median, the distribution is likely **right-skewed**. When mean < median, it is **left-skewed**.

---

## 3. Visualizing Distributions

### 3.1 Histogram

A histogram groups continuous values into bins and shows their frequency.

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Highway MPG",
    x = "Highway Miles Per Gallon",
    y = "Count"
  ) +
  theme_minimal()
```

**What to look for:**

- Is the distribution symmetric or skewed?
- Are there multiple peaks (bimodal)?
- Are there gaps or extreme values?

### 3.2 Density Plot

A density plot is a smoothed version of the histogram, useful for comparing groups.

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_density(fill = "steelblue", alpha = 0.5) +
  labs(
    title = "Density of Highway MPG",
    x = "Highway MPG",
    y = "Density"
  ) +
  theme_minimal()
```

### 3.3 Boxplot

Boxplots show the five-number summary: minimum, Q1, median, Q3, maximum — and flag potential outliers.

```{r}
ggplot(mpg, aes(x = class, y = hwy, fill = class)) +
  geom_boxplot(show.legend = FALSE) +
  labs(
    title = "Highway MPG by Vehicle Class",
    x = "Class",
    y = "Highway MPG"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## 4. Outlier Detection

Outliers can be legitimate extreme values or data entry errors. A common rule is the **IQR method**:

- Lower fence: Q1 − 1.5 × IQR
- Upper fence: Q3 + 1.5 × IQR

```{r}
Q1 <- quantile(mpg$hwy, 0.25)
Q3 <- quantile(mpg$hwy, 0.75)
IQR_val <- Q3 - Q1

lower <- Q1 - 1.5 * IQR_val
upper <- Q3 + 1.5 * IQR_val

outliers <- mpg %>% filter(hwy < lower | hwy > upper)
cat("Number of outliers:", nrow(outliers), "\n")
outliers
```

> **Important:** Outliers should not be removed automatically. Always investigate whether they represent real observations or data errors.

---

## 5. Data Transformation

When data is strongly skewed, transformations can help normalize it for analysis and modeling.

### 5.1 Log Transformation

```{r}
# Simulate right-skewed data
set.seed(42)
skewed_data <- tibble(x = exp(rnorm(500, mean = 3, sd = 1)))

# Original vs log-transformed
p1 <- ggplot(skewed_data, aes(x = x)) +
  geom_histogram(fill = "salmon", color = "white", bins = 30) +
  labs(title = "Original (Right-Skewed)", x = "x") +
  theme_minimal()

p2 <- ggplot(skewed_data, aes(x = log(x))) +
  geom_histogram(fill = "steelblue", color = "white", bins = 30) +
  labs(title = "Log-Transformed", x = "log(x)") +
  theme_minimal()

library(patchwork)
p1 + p2
```

---

## 6. Grouped Summaries with `dplyr`

A key part of EDA is comparing distributions across groups.

```{r}
mpg %>%
  group_by(class) %>%
  summarise(
    n          = n(),
    mean_hwy   = mean(hwy),
    sd_hwy     = sd(hwy),
    median_hwy = median(hwy)
  ) %>%
  arrange(desc(mean_hwy))
```

```{r}
# Visualize group means
mpg %>%
  group_by(class) %>%
  summarise(mean_hwy = mean(hwy)) %>%
  ggplot(aes(x = reorder(class, mean_hwy), y = mean_hwy, fill = class)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Average Highway MPG by Vehicle Class",
    x = "Class",
    y = "Mean Highway MPG"
  ) +
  theme_minimal()
```

---

## 7. Summary

| Concept              | Tool                              |
|----------------------|-----------------------------------|
| Distribution shape   | Histogram, Density plot           |
| Center & spread      | `mean()`, `median()`, `sd()`, `IQR()` |
| Outlier detection    | Boxplot, IQR rule                 |
| Skewness correction  | Log / sqrt transformation         |
| Group comparison     | `group_by()` + `summarise()`      |

---

## Key Takeaways

- Always visualize distributions before computing statistics.
- The mean is sensitive to outliers; the median is more robust.
- Use boxplots to compare groups and spot outliers simultaneously.
- Log transformation is effective for right-skewed data.
- Grouped summaries reveal patterns that overall summaries hide.

