---
title: "week 03 - Working with Files and Data in R"
format: html
---



```{r}
# Create a matrix
m <- matrix(1:9, nrow = 3)
m
```

my goal is to calculate the mean of each row of this matrix


```{r}
mean(c(1,4,7))
mean(c(2,5,8))
mean(c(3,6,9))
```

loops


```{r}
# Calculate row means using a for-loop
row_means <- numeric(nrow(m))
for (i in 1:nrow(m)) {
  row_means[i] <- mean(m[i, ])
}
row_means
```


```{r}
# Cleaner and more direct
apply(m, 1, mean)
```

## lapply

loop first


```{r}
data_list <- list(a = 1:5, b = 6:10, c = 11:15)
squared_list <- list()

for (i in seq_along(data_list)) {
  squared_list[[i]] <- data_list[[i]]^2
}
squared_list
```

```{r}
y <- function(x) {sin(x)}
y(3.1415)
```


```{r}
lapply(data_list, function(x) x^2)
```

## sapply


```{r}
# Returns a named numeric vector instead of a list
sapply(data_list, mean)
```

Task for you what is the structure of this output? It is a vector, a list or something else?

comparison with a loop


```{r}
means <- numeric(length(data_list))
for (i in seq_along(data_list)) {
  means[i] <- mean(data_list[[i]])
}
means
```

## vapply


```{r}
vapply(data_list, mean, numeric(1))
```

## tapply()


```{r}
heights <- c(160, 170, 180, 150, 165, 175)
gender <- factor(c("F", "M", "M", "F", "F", "M"))

# Average height by gender
tapply(heights, gender, mean)
```

loop version

```{r}
means <- numeric(length(levels(gender)))
for (i in seq_along(levels(gender))) {
  group <- levels(gender)[i]
  means[i] <- mean(heights[gender == group])
}
means
```


## mapply

apply a function to multiple inputs in parallel


```{r}
# Multiply corresponding elements
x <- 1:5
y <- 6:10

mapply(function(a, b) a * b, x, y)

x*y
```

### Friday: Datasets in r

### 1. Built-in datasets in base R

```{r}
data()
```

```{r}
data(mtcars)
?mtcars

df <- mtcars
str(df)
head(df)[10]
tail(df)[10]
```

```{r}
data(penguins , package = "palmerpenguins")
```

```{r}
df <- penguins
str(df)

```

Statistical Summaries


```{r}
summary(df)
```

## Exercise 

# List all available datasets in R

```{r}
data()
```
# Load the iris dataset

```{r}
df <- iris
```
# Inspect the structure of the dataset

```{r}
str(df)
```
# View a statistical summary of the dataset
```{r}
summary(df)
```


```{r}
?iris
```

## How to find information

```{r}
help.search("motor")
```


## Exercise 2

```{r}
help(iris)
```

what the measurements represent:

Whether the dataset includes a target or class variable?


## identify dataset from what package

```{r}
data(package = "datasets")
```



```{r}
data(package = "palmerpenguins")
data(package = "modeldata")
```


## Exercise 3

Load one dataset from {palmerpenguins} or {modeldata}.

Print:

dim()
names()
str()



## Example: tidy inspection with tidyverse

```{r}
library(tidyverse)

penguins %>%
select(species, island, bill_length_mm, body_mass_g) %>%
drop_na() %>%
summary()
```