---
title: "Week 04 - Exploratory Data Analysis (EDA)"
format: html
---

## Monday

## Exploratory Data Analysis (EDA)




## EDA Mindset

Is iterative, investigative and visual.

Expect:
-Go back and forth between steps 
-Revise assumptions
-Discover data issues late in the process
-Ask better questions as you learn more



## Understanding the context

Before touching the data understand:
-Where did the data come from
-Why was it collected
- Who or what does each row represent?
-What decisions will this data support?

Key questions
-Is this observational or experimental data?
-Are there known biases?
-Is time involved
-Are there units, definitions, or domain rules?

Poor context understanding leads to correct analysis of the wrong problem.



## Data Validation (Trust but verify)

Validation answers one core question:

Is this data structurally and logically valid?

This step focuses on correctness, not insight.



## Structural Validation

-Do all columns exist?
-Are data types correct
-Are there unexpected nulls?
-Are identifieres unique when they should be?



## Logical Validation

-Are values within expected ranges?
-Do dates make sense chronologically?
-Do relationships between variables violate domain rules?

RED FLAGS:

Negative ages
Future timestamps
Duplicate IDs
Impossible combination of values



## Data Cleaning(Fixing What's Broken)

Cleaning addresses errors, inconsistencies, and noise.

Typical tasks:

-Handling missing values 
-Fixing inconsistent labels or categories
-Correcting obvious data entry errors 
-Removing or flagging duplicates

Important principle:

Cleaning should be conservative and documented.

Avoid "fixing" data unless:

-You know the cause
-The correction is defensible
-you preserve original values when possible



## Data Massaging (Making Data Usable)

Massaging prepares data for analysis and exploration, not modeling.

This step is about shape and representation.

Common activities:

-Renaming variables for clarity
-Creating derived variables
-Converting units or scales
-Reshaping data (wide - Long)
-Standardizing formats (dates, categories, text)

Key idea:
    Clean data can still be unusable massaging makes it analysis-ready.



## Univariate Exploration (One Variable at a Time)

Now you start learning what the data looks like.

For each variable, ask:

-What values occur?
-What is typical?
-What is extreme?
-What is missing?

Conceptually

Numeric variables distribution, spread, skewness

Categorical variables frequency, dominance, rarity

This step builds baseline intuition.



## Bivariate Exploration (Relationships)

Next, explore how variables relate to each other.

Ask questions like:

-Does one variable change with another?
-Are there strong associations?
-Are relationships linear, nonlinear, or absent?

Focus on:

Direction
Strength
Consistency
Exceptions

    Many insights live in relationships, not individual variables.



## Multivariate Exploration (Context Matters)

Multivariate analysis looks at relationships conditiones on other variables.

This helps answer:

Does this relationship hold across groups?
Is this pattern driven by a subset?
Are there interaction effects?

This step often reveals:

Hidden structure 
Cofounding Variables
Simpson's paradox-type effects



## Outliers and Anomalies (Signal or Noise?)

Outliers are not automatically bad.

Ask:
Are they errors?
Are they rare but valid cases?
Do they represent a different process?

Important distinction:
Outlier extreme but valid
Anomaly inconsistent with data-generating process

    Never remove outliers without understanding them.



## Assumptions checking

Before modeling, EDA helps you check:

Distributional assumptions
Independence
Stationarity (for time-based data)
Linearity or monotonicity

This step informs:

Feature engineering
Model Choice
Transformation needs



## EDA is Iterative

EDA does not end cleanly.

You will:

Discover issues late
Revisit Cleaning
Create new variables mid-analysis
change direction based on findings

This is normal and expected.



## Key Takeaways

EDA is about understanding, not proving

Most errors are found before modeling 



## Wednesday (Lab EDA)


```{r}
?airquality
```


## Libraries to install



```{r}
# if you don't have these installed yet, run:
# install.packages(c("dplyr", "ggplot2" , "tidyr" , "skimr"))
library(dplyr)
library(skimr)
library(ggplot2)
library(tidyr)


```

## 1.Explore

 Load airquality to a new variable

```{r}
aq_raw <- airquality
```




## Lets eliminate the rows which have missing values for ozone

```{r}
library(dplyr)

aq_raw |>
    filter(
    is.na(Ozone)
    )
```


### Let's use skimr


```{r}
skim(aq_raw)
```

```{r}
aq_little_clean <- aq_raw |>
  filter(!is.na(Ozone))
```

```{r}
skim(aq_little_clean)
```


## 2. Validate the data



```{r}
nrow(aq_raw)
nrow(aq_little_clean)
```



```{r}
names(aq_raw)
```

Total NA's per variables

```{r}
colSums(is.na(aq_raw))

```

Script logical validation to check if any values is suspecious

```{r}
aq_raw %>%
summarize(
month_min = min(Month, na.rm = TRUE),
month_max = max(Month, na.rm = TRUE),
day_min   = min(Day, na.rm = TRUE),
day_max   = max(Day, na.rm = TRUE),
temp_min  = min(Temp, na.rm = TRUE),
temp_max  = max(Temp, na.rm = TRUE),
wind_min  = min(Wind, na.rm = TRUE),
wind_max  = max(Wind, na.rm = TRUE)
)
```


## 3.Cleaning

**Checking missing**


```{r}
aq_raw %>%
summarize(
missing_ozone  = sum(is.na(Ozone)),
missing_solarR = sum(is.na(Solar.R)),
missing_wind   = sum(is.na(Wind)),
missing_temp   = sum(is.na(Temp))
)
```

### 3.2 Decide on a missing data strategy


```{r}
aq_clean <- aq_raw %>%
group_by(Month) %>%
mutate(
Ozone   = ifelse(is.na(Ozone),   median(Ozone, na.rm = TRUE),   Ozone),
Solar.R = ifelse(is.na(Solar.R), median(Solar.R, na.rm = TRUE), Solar.R)
) %>%
ungroup()

colSums(is.na(aq_clean))
```

## Do i have duplicates?



```{r}
aq_clean %>%
count(Month, Day) %>%
filter(n > 1)
```


### 4.Massage the data (make it analysis-ready)



```{r}
aq_tidy <- aq_clean %>%
rename(
ozone = Ozone,
solar = Solar.R,
wind  = Wind,
temp  = Temp,
month = Month,
day   = Day
)

names(aq_tidy)
```


### 4.2 Create a proper date column



```{r}
aq_tidy <- aq_tidy %>%
mutate(date = as.Date(sprintf("1973-%02d-%02d", month, day)))

min(aq_tidy$date); max(aq_tidy$date)
```

### Changing the type of Month

from int to factor



```{r}
aq_tidy <- aq_tidy %>%
mutate(
month_name = factor(
month,
levels = 5:9,
labels = c("May", "Jun", "Jul", "Aug", "Sep")
)
)

table(aq_tidy$month_name)
```

## Share your data with others



```{r}
aq_tidy <- aq_tidy %>%
mutate(
temp_c = (temp - 32) * 5/9,
high_ozone = ozone > 84
)

aq_tidy %>% select(date, ozone, high_ozone, temp, temp_c) %>% head()
```

## 5.Exploratory Data Analysis (EDA)

### Univariate Exploration

```{r}
aq_tidy %>%
summarize(
ozone_mean = mean(ozone),
ozone_median = median(ozone),
ozone_sd = sd(ozone),
temp_mean = mean(temp),
wind_mean = mean(wind)
)
```

### Let's use hist() first

```{r}
hist(aq_tidy$ozone)
```

### Let's use ggplot for distribution visualization

```{r}
ggplot(aq_tidy, aes(x = ozone)) +
    geom_histogram(bins = 20) +
    labs(title = "Distribution of Ozone", x = "Ozone", y = "Count")
```
