---
title: "Week 07 - Bivariate and Multivariate Analysis"
format: html
---

## Overview

This week we move beyond single-variable summaries into **bivariate** (two variables) and **multivariate** (three or more variables) analysis. Understanding relationships between variables is the foundation of data-driven insight and predictive modeling.

By the end of this week you should be able to:

- Compute and interpret correlation coefficients
- Visualize relationships between two continuous variables
- Analyze relationships involving categorical variables
- Build and interpret a correlation matrix
- Visualize three or more variables simultaneously

---

## 1. Bivariate Analysis

Bivariate analysis examines the relationship between **two variables**. The approach depends on the variable types:

| Variable 1   | Variable 2   | Method                          |
|--------------|--------------|---------------------------------|
| Continuous   | Continuous   | Scatter plot, correlation       |
| Continuous   | Categorical  | Boxplot, grouped density        |
| Categorical  | Categorical  | Bar chart, contingency table    |

---

## 2. Continuous vs. Continuous

### 2.1 Scatter Plot

The scatter plot is the most direct way to visualize a relationship between two numeric variables.

```{r}
library(tidyverse)
data(mpg)

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5, color = "steelblue") +
  geom_smooth(method = "lm", se = TRUE, color = "tomato") +
  labs(
    title = "Engine Displacement vs Highway MPG",
    x = "Engine Displacement (L)",
    y = "Highway MPG"
  ) +
  theme_minimal()
```

### 2.2 Pearson Correlation

The **Pearson correlation coefficient (r)** measures the strength and direction of a linear relationship. It ranges from −1 to +1:

- **r = 1**: perfect positive linear relationship
- **r = 0**: no linear relationship
- **r = −1**: perfect negative linear relationship

```{r}
cor_val <- cor(mpg$displ, mpg$hwy)
cat("Pearson r:", round(cor_val, 3), "\n")
```

```{r}
# Correlation test with p-value
cor.test(mpg$displ, mpg$hwy)
```

> **Interpretation:** A negative correlation confirms that larger engines tend to have worse fuel efficiency. A p-value below 0.05 indicates this relationship is statistically significant.

> **Caution:** Correlation measures **linear** association only. Always pair it with a scatter plot — two variables can have a strong non-linear relationship but a near-zero Pearson r.

---

## 3. Continuous vs. Categorical

### 3.1 Grouped Boxplot

```{r}
ggplot(mpg, aes(x = reorder(class, hwy, median), y = hwy, fill = class)) +
  geom_boxplot(show.legend = FALSE) +
  coord_flip() +
  labs(
    title = "Highway MPG by Vehicle Class",
    x = "Class",
    y = "Highway MPG"
  ) +
  theme_minimal()
```

### 3.2 Grouped Density Plot

```{r}
ggplot(mpg, aes(x = hwy, fill = drv)) +
  geom_density(alpha = 0.5) +
  labs(
    title = "Highway MPG Distribution by Drive Type",
    x = "Highway MPG",
    y = "Density",
    fill = "Drive Type"
  ) +
  theme_minimal()
```

### 3.3 Violin Plot

A violin plot combines a boxplot with a density estimate to show the full distribution shape.

```{r}
ggplot(mpg, aes(x = drv, y = hwy, fill = drv)) +
  geom_violin(trim = FALSE, show.legend = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", show.legend = FALSE) +
  labs(
    title = "Highway MPG by Drive Type (Violin + Boxplot)",
    x = "Drive Type",
    y = "Highway MPG"
  ) +
  theme_minimal()
```

---

## 4. Categorical vs. Categorical

### 4.1 Contingency Table

```{r}
table(mpg$drv, mpg$class)
```

### 4.2 Stacked and Grouped Bar Charts

```{r}
# Proportional stacked bar
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar(position = "fill") +
  labs(
    title = "Drive Type Proportion within Each Vehicle Class",
    x = "Class",
    y = "Proportion",
    fill = "Drive Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

```{r}
# Grouped bar chart
ggplot(mpg, aes(x = class, fill = drv)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Count of Drive Types by Vehicle Class",
    x = "Class",
    y = "Count",
    fill = "Drive Type"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## 5. Correlation Matrix

When working with multiple continuous variables, a correlation matrix gives a complete overview of pairwise relationships.

```{r}
# Select numeric columns
num_vars <- mpg %>% select(displ, cyl, cty, hwy)

# Compute correlation matrix
cor_matrix <- cor(num_vars)
round(cor_matrix, 2)
```

### 5.1 Visualizing with a Heatmap

```{r}
library(reshape2)

# Melt correlation matrix for ggplot
cor_melted <- melt(cor_matrix)

ggplot(cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(value, 2)), size = 4) +
  scale_fill_gradient2(
    low = "steelblue", mid = "white", high = "tomato",
    midpoint = 0, limits = c(-1, 1)
  ) +
  labs(
    title = "Correlation Matrix Heatmap",
    x = NULL, y = NULL, fill = "r"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## 6. Multivariate Visualization

### 6.1 Encoding a Third Variable with Color or Size

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class, size = cyl)) +
  geom_point(alpha = 0.6) +
  labs(
    title = "Displacement vs MPG (Color = Class, Size = Cylinders)",
    x = "Displacement (L)",
    y = "Highway MPG",
    color = "Class",
    size = "Cylinders"
  ) +
  theme_minimal()
```

### 6.2 Faceted Scatter Plots

Facets allow you to visualize a relationship across subgroups.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ year) +
  labs(
    title = "Displacement vs MPG by Drive Type, Faceted by Year",
    x = "Displacement (L)",
    y = "Highway MPG",
    color = "Drive Type"
  ) +
  theme_minimal()
```

### 6.3 Scatterplot Matrix with `GGally`

A scatterplot matrix shows all pairwise relationships at once — great for an initial multivariate overview.

```{r}
# install.packages("GGally") if needed
library(GGally)

mpg %>%
  select(displ, cyl, cty, hwy, class) %>%
  ggpairs(
    aes(color = class, alpha = 0.5),
    upper = list(continuous = wrap("cor", size = 3)),
    lower = list(continuous = wrap("points", alpha = 0.3))
  ) +
  theme_minimal() +
  labs(title = "Scatterplot Matrix of Key MPG Variables")
```

---

## 7. Summary Table

| Relationship Type      | Visualization            | Statistic               |
|------------------------|--------------------------|-------------------------|
| Continuous–Continuous  | Scatter + smooth line    | Pearson r               |
| Continuous–Categorical | Boxplot, Violin, Density | Group means / medians   |
| Categorical–Categorical| Stacked/grouped bar      | Contingency table       |
| Multi-variable         | Facets, color+size, GGally scatterplot matrix | Correlation matrix |

---

## Key Takeaways

- Always visualize relationships before computing statistics.
- Pearson r only captures **linear** relationships — check with a scatter plot.
- Boxplots and violin plots are powerful for comparing a numeric variable across groups.
- Use proportional bar charts (`position = "fill"`) when comparing categorical distributions of unequal group sizes.
- Correlation matrices and scatterplot matrices (`ggpairs`) are efficient for exploring many variables at once.
- Facets are your best tool for adding a third variable to a bivariate plot without cluttering it.

