---
title: "Week 06 - ggplot"
format: html
---

## Overview

This week we take a deep dive into **ggplot2**, R's most powerful and flexible visualization library. ggplot2 is built on the **Grammar of Graphics**, a principled framework for building plots layer by layer.

By the end of this week you should be able to:

- Understand the grammar of graphics and the `ggplot()` structure
- Map variables to aesthetics (color, size, shape, etc.)
- Use a variety of `geom_*` functions
- Customize themes, labels, scales, and facets
- Combine multiple plots

---

## 1. The Grammar of Graphics

Every ggplot2 plot is built from the same components:

| Component   | Description                                      |
|-------------|--------------------------------------------------|
| **Data**    | The dataset being visualized                     |
| **Aesthetics** (`aes`) | Mappings from variables to visual properties |
| **Geoms**   | The geometric objects drawn (points, bars, lines)|
| **Stats**   | Statistical transformations (counts, smoothing)  |
| **Scales**  | Controls over axes, colors, sizes                |
| **Facets**  | Splitting the plot into subplots                 |
| **Theme**   | Non-data visual elements (fonts, background)     |

The basic syntax is always:

```r
ggplot(data = <DATA>, aes(<MAPPINGS>)) +
  geom_<TYPE>(<OPTIONS>)
```

---

## 2. Setting Up

```{r}
library(tidyverse)

# We'll use mpg and diamonds datasets throughout
data(mpg)
data(diamonds)
```

---

## 3. Core Geoms

### 3.1 Scatter Plot — `geom_point()`

Used to visualize the relationship between two continuous variables.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue", alpha = 0.6, size = 2) +
  labs(
    title = "Engine Displacement vs Highway MPG",
    x = "Engine Displacement (L)",
    y = "Highway MPG"
  ) +
  theme_minimal()
```

### 3.2 Adding Color by Group

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Engine Displacement vs Highway MPG by Class",
    x = "Engine Displacement (L)",
    y = "Highway MPG",
    color = "Vehicle Class"
  ) +
  theme_minimal()
```

### 3.3 Line Plot — `geom_line()`

Useful for time series or ordered data.

```{r}
# Average MPG by year and drive type
mpg %>%
  group_by(year, drv) %>%
  summarise(mean_hwy = mean(hwy), .groups = "drop") %>%
  ggplot(aes(x = year, y = mean_hwy, color = drv, group = drv)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  labs(
    title = "Average Highway MPG Over Time by Drive Type",
    x = "Year",
    y = "Mean Highway MPG",
    color = "Drive Type"
  ) +
  theme_minimal()
```

### 3.4 Bar Chart — `geom_bar()` and `geom_col()`

`geom_bar()` counts observations. `geom_col()` uses a pre-computed value.

```{r}
# Count of cars by class
ggplot(mpg, aes(x = fct_infreq(class), fill = class)) +
  geom_bar(show.legend = FALSE) +
  labs(
    title = "Number of Vehicles by Class",
    x = "Vehicle Class",
    y = "Count"
  ) +
  theme_minimal()
```

```{r}
# Pre-computed averages with geom_col
mpg %>%
  group_by(class) %>%
  summarise(mean_hwy = mean(hwy)) %>%
  ggplot(aes(x = reorder(class, mean_hwy), y = mean_hwy, fill = class)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Average Highway MPG by Class", x = "Class", y = "Mean MPG") +
  theme_minimal()
```

### 3.5 Histogram — `geom_histogram()`

```{r}
ggplot(mpg, aes(x = hwy)) +
  geom_histogram(bins = 20, fill = "steelblue", color = "white") +
  labs(title = "Highway MPG Distribution", x = "Highway MPG", y = "Count") +
  theme_minimal()
```

### 3.6 Boxplot — `geom_boxplot()`

```{r}
ggplot(mpg, aes(x = drv, y = hwy, fill = drv)) +
  geom_boxplot(show.legend = FALSE) +
  labs(title = "Highway MPG by Drive Type", x = "Drive Type", y = "Highway MPG") +
  theme_minimal()
```

### 3.7 Smooth Line — `geom_smooth()`

Adds a regression line or LOESS smoother.

```{r}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4, color = "gray50") +
  geom_smooth(method = "lm", color = "tomato", se = TRUE) +
  labs(
    title = "Engine Displacement vs Highway MPG (with Linear Fit)",
    x = "Displacement (L)",
    y = "Highway MPG"
  ) +
  theme_minimal()
```

---

## 4. Aesthetics in Depth

Aesthetics can be **mapped** (tied to a variable) or **set** (fixed for all observations).

```{r}
# Mapped: color varies by class
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()

# Set: all points are red
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "red")
```

Common aesthetics: `color`, `fill`, `size`, `shape`, `alpha`, `linetype`.

---

## 5. Scales

Scales control how aesthetic mappings look.

```{r}
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set2") +
  scale_x_continuous(breaks = seq(1, 7, by = 1)) +
  labs(title = "Custom Color Scale and Axis Breaks") +
  theme_minimal()
```

---

## 6. Facets

Facets split a plot into subplots based on one or two variables.

```{r}
# facet_wrap: one variable
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "steelblue", alpha = 0.6) +
  facet_wrap(~ class, ncol = 3) +
  labs(title = "Displacement vs MPG Faceted by Class") +
  theme_minimal()
```

```{r}
# facet_grid: two variables
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.5) +
  facet_grid(drv ~ cyl) +
  labs(title = "Faceted by Drive Type (rows) and Cylinders (cols)") +
  theme_minimal()
```

---

## 7. Themes and Labels

```{r}
ggplot(mpg, aes(x = class, y = hwy, fill = class)) +
  geom_violin(show.legend = FALSE, trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", show.legend = FALSE) +
  labs(
    title    = "Highway MPG Distribution by Vehicle Class",
    subtitle = "Violin plot with embedded boxplot",
    caption  = "Source: ggplot2::mpg dataset",
    x = "Vehicle Class",
    y = "Highway MPG"
  ) +
  theme_classic() +
  theme(
    plot.title    = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(color = "gray40"),
    axis.text.x   = element_text(angle = 30, hjust = 1)
  )
```

**Built-in themes:** `theme_minimal()`, `theme_classic()`, `theme_bw()`, `theme_light()`, `theme_dark()`.

---

## 8. Combining Plots with `patchwork`

```{r}
library(patchwork)

p1 <- ggplot(mpg, aes(x = hwy)) +
  geom_histogram(fill = "steelblue", bins = 20) +
  labs(title = "Histogram") + theme_minimal()

p2 <- ggplot(mpg, aes(x = class, y = hwy)) +
  geom_boxplot(fill = "salmon") +
  labs(title = "Boxplot") + theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p1 + p2
```

---

## Key Takeaways

- ggplot2 builds plots in layers using the Grammar of Graphics.
- Always start with `ggplot(data, aes(...))` then add geoms.
- Aesthetic mappings link variables to visual properties; setting fixes them.
- Use `facet_wrap()` or `facet_grid()` to create small multiples.
- `theme_*()` functions quickly change the overall look of a plot.
- `patchwork` makes combining multiple plots easy.

