---
title: "Week 02 - R Refresher"
format: html
---

## How we include figures in quarto?

Here we will explore the basis and main data structures in R, for example

### Vectors
```{r}
my_little_pony <- c("Twilight Sparkle", "Rainbow Dash", "Pinkie Pie", "Applejack", "Rarity")
numbers_Vector <- c(seq(2,30,by=2))
x_log <- c(TRUE, FALSE, FALSE, FALSE, TRUE)
```

### Matrices
```{r}
A <- matrix(1:6, nrow = 2, ncol = 3, byrow = TRUE)
print(A)
A
```

### 3D array
```{r}
a <- array(1:27, dim = c(3, 3, 3))
a
```

### Lists
```{r}
my_list <- list(
  numbers = 1:3,
  letters = c("a", "b"),
  matrix  = matrix(1:4, nrow = 2)
)
my_list
```
```{r}
my_list$letters
```

### DataFrames
```{r}
vector1 <- c(seq(2,6,1))
vector2 <- c(seq(3,7,1))
vector3 <- c(letters[1:5])
df <- data.frame(vector1, vector2)
df
df$vectors4 <- c(TRUE, FALSE, FALSE, FALSE, TRUE)
```
```{r}
id    <- 1:5
name  <- c("Alice", "Bob", "Cara", "Dan", "Enrique")
score <- c(88, 92, 79, 85, NA)

students <- data.frame(id, name, score)
students

mean(students$score, na.rm = TRUE)
str(students)
is.na(students$score)
```

### Add a new column using a condition
```{r}
students$passed <- students$score >= 85
students
```
```{r}
students$grade <- ifelse(
  students$score >= 90, "A",
  ifelse(students$score >= 80, "B", "C")
)
students
```

### Renaming variable names from a dataframe
```{r}
names(students)
```

### Selecting using square brackets
```{r}
students[, "score"]
students[, 3]
students[, c("score", "name")]
students[, -1]
```

### Subselecting from a dataframe
```{r}
df <- students[students$score > 80, ]
df
remove(df)
```

### Factors
```{r}
gender_char <- c("F", "M", "F", "F", "M")
gender_fact <- factor(gender_char)
gender_fact
```

### Directory Paths
```{r}
getwd()
```

### Importing Data — CSV with readr
```{r}
library(readr)

# Simulated data (in practice: students <- read_csv("data/students.csv"))
students <- data.frame(
  id    = 1:5,
  name  = c("Alice", "Bob", "Cara", "Dan", "Enrique"),
  score = c(88, 92, 79, 85, 91)
)
students
```

### Excel with readxl
```{r}
library(readxl)

# Simulated data (in practice: grades <- read_excel("data/grades.xlsx"))
grades <- data.frame(
  id      = 1:4,
  znumber = c("Z001", "Z002", "Z003", "Z004"),
  midterm = c(85, 90, 78, 88),
  final   = c(88, 92, 80, 85)
)
grades
```

### Parquet with arrow
```{r}
library(arrow)

# Simulated data (in practice: big_data <- read_parquet("data/big_data.parquet"))
big_data <- data.frame(
  id    = 1:6,
  value = c(10.5, 20.1, 30.3, 40.7, 50.2, 60.9),
  group = c("A", "A", "B", "B", "C", "C")
)
big_data
```

### RData and RDS
```{r}
# Simulating save/load with temp files
simple_model <- lm(score ~ id, data = students)
tmp <- tempfile(fileext = ".rds")
saveRDS(simple_model, tmp)
model_restored <- readRDS(tmp)
summary(model_restored)
```

### Writing CSV
```{r}
library(readr)
write_csv(grades, "grades_clean.csv")
cat("grades_clean.csv written successfully\n")
```

### Writing Excel
```{r}
library(writexl)
write_xlsx(grades, "grades_export.xlsx")
cat("grades_export.xlsx written successfully\n")
```

### Writing Parquet
```{r}
library(arrow)
write_parquet(big_data, "big_data.parquet")
cat("big_data.parquet written successfully\n")
```

### Saving RData and RDS
```{r}
save(students, grades, file = "class_data.RData")
saveRDS(students, file = "students_clean.rds")
load("class_data.RData")
students_restored <- readRDS("students_clean.rds")
students_restored
```
